Expressive Array Constructs in an Embedded GPU Kernel Programming Language

### 2013-10-06 src2/embarbb/results.tex
apa
hacking

\subsection{Benchmarks}
\label{sec:Benchmarks}
%\FloatBarrier

In this section, matrix multiply, Sobel edge detection and an image blur
algorithm are used as benchmarks in comparing sequential C++ code, ArBB, EmbArBB, 
and a Haskell library called Repa. 
Repa provides regular shape polymorphic arrays; it permits
parallel execution of the resulting code, making use of multiple
cores, but not of SIMD parallelism~\cite{REPA}.
The Repa versions of the benchmarks used in the comparison 
come from the {\tt repa-examples-3.2.1.1} package on Hackage.

>>>>

\subsection{Benchmarks}
\label{sec:EmbArBBBenchmarks}
%\FloatBarrier

In this section, matrix multiply, Sobel edge detection and an image blur
algorithm are used as benchmarks in comparing sequential C++ code, ArBB, EmbArBB, 
and a Haskell library called Repa. 
Repa provides regular shape polymorphic arrays; it permits
parallel execution of the resulting code, making use of multiple
cores, but not of SIMD parallelism~\citeemb{REPA}.
The Repa versions of the benchmarks used in the comparison 
come from the {\tt repa-examples-3.2.1.1} package on Hackage.

### 2013-10-06 src2/expressive/applications.tex
apa
hacking

Sorting kernels are typically used as building blocks in larger
programs to sort much larger sequences of inputs. In section~\ref{sec:benchmarks},
we show how to build a sorter for large arrays from small building blocks, including
small kernels for sorting and merging that are generated from Obsidian.
Our small kernels are constructed in the form of sorting and merging {\em networks},
building on Batcher's bitonic merger~\cite{Batcher} and
on the periodic balanced merger~\cite{PeriodicBalanced}.
We chose also to implement the large sorter used  in benchmarking the
small kernels as a sorting network. However, large
sorters that are not themselves sorting networks (with typical examples being radix sort and quicksort) often call small sorting networks when they need to sort small arrays during their execution.
Thus, small, fast sorting kernels have a variety of uses.

@@

The bitonic merger is typically presented as a recursive
construction and we have earlier explored ways to describe
and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LAVASORTER}.
Here, we consider iterative descriptions using similar combinators.

@@

It is in combining the results of the {\tt f} and {\tt g} functions
that we run into difficulty using just Pull arrays.
Earlier, we saw how to use Push arrays to implement {\tt concP},
which concatenates two arrays.
Here, we use exactly the same approach to make a new version of
the {\em interleave} combinator. The results of applying
the {\tt f}s and {\tt gs} are combined into a Push array, in the right
order.
\pagebreak
%% ** TODO More explanation needed.
\begin{codesize}
\begin{verbatim}
ixMap :: (UWordE -> UWordE) -> ArrayP a -> ArrayP a 
ixMap f (ArrayP p n) = ArrayP (ixMap' f p) n

@@

In order to go faster still, we resort to building a different sorting network, of
exactly the same size as the bitonic sorter, but based instead on the balanced period merger of Dowd et al~\cite{PeriodicBalanced}.
This involves the introduction of one new combinator that can be seen as a mixture
of the {\tt ilv} and {\tt vee} combinators already introduced.

@@

    bmergeSmall<<<blocks,threads,4096>>>(din,din);}
\end{verbatim}
\end{codesize}
\caption{CUDA code for our large sorter. {\tt sortSmall} and {\tt bmergeSmall}
are replaced by Obsidian-generated kernels in the experiments. {\tt vSwap} and {\tt iSwap} are handwritten CUDA kernels
that perform one column of compare and swap operations in the vee and ilv shapes respectively, and are
parameterised on the stride. (Our generated kernels have fixed input and output size.)}
\label{fig:CUDAsort}
\end{figure}
We have measured the performance of generated 512-input sorting and merging kernels by plugging them into a larger sorter written in CUDA.
The sorter has exactly the structure shown in Figure~\ref{fig:mixedsorter}.
Figure~\ref{fig:newmixedsorter} shows the location of smaller sorters and mergers, and of vSwap and iSwap kernels for $16$ inputs.
Larger sorters simply have more columns of mergers, each preceded by a vSwap and a number of iSwaps.
The overall structure of the resulting CUDA code is shown in Figure~\ref{fig:CUDAsort}.
Because {\tt iSwap} is used repeatedly, we also wrote kernels
corresponding to compositions of two and three of them (avoiding
memory accesses between the columns).
That is, we replaced the loop containing {\tt iSwap} above by
\pagebreak
\begin{codesize}
\begin{verbatim}
for(int j = i-1; j >= 0; j -= 3){ 
  if (j==0) 
    iSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
    else 
    {if (j==1) 
       iSwap2<<<blocks/4,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
       else 
       iSwap3<<<blocks/8,threads*2,0>>>(din,din,(1<<j)*S_SIZE);}}
\end{verbatim}
\end{codesize}
\noindent
The Obsidian implementation and all code for examples in the paper are available at \url{http://www.cse.chalmers.se/~joels/expressive.html}.

>>>>

Sorting kernels are typically used as building blocks in larger
programs to sort much larger sequences of inputs. In section~\ref{sec:benchmarks},
we show how to build a sorter for large arrays from small building blocks, including
small kernels for sorting and merging that are generated from Obsidian.
Our small kernels are constructed in the form of sorting and merging {\em networks},
building on Batcher's bitonic merger~\citeexp{Batcher} and
on the periodic balanced merger~\citeexp{PeriodicBalanced}.
We chose also to implement the large sorter used  in benchmarking the
small kernels as a sorting network. However, large
sorters that are not themselves sorting networks (with typical examples being radix sort and quicksort) often call small sorting networks when they need to sort small arrays during their execution.
Thus, small, fast sorting kernels have a variety of uses.

@@

The bitonic merger is typically presented as a recursive
construction and we have earlier explored ways to describe
and analyse it in both (our) Ruby and in Lava~\citeexp{sortsRuby,LAVASORTER}.
Here, we consider iterative descriptions using similar combinators.

@@

It is in combining the results of the {\tt f} and {\tt g} functions
that we run into difficulty using just Pull arrays.
Earlier, we saw how to use Push arrays to implement {\tt concP},
which concatenates two arrays.
Here, we use exactly the same approach to make a new version of
the {\em interleave} combinator. The results of applying
the {\tt f}s and {\tt gs} are combined into a Push array, in the right
order.
%\pagebreak
%% ** TODO More explanation needed.
\begin{codesize}
\begin{verbatim}
ixMap :: (UWordE -> UWordE) -> ArrayP a -> ArrayP a 
ixMap f (ArrayP p n) = ArrayP (ixMap' f p) n

@@

In order to go faster still, we resort to building a different sorting network, of
exactly the same size as the bitonic sorter, but based instead on the balanced period merger of Dowd et al~\citeexp{PeriodicBalanced}.
This involves the introduction of one new combinator that can be seen as a mixture
of the {\tt ilv} and {\tt vee} combinators already introduced.

@@

    bmergeSmall<<<blocks,threads,4096>>>(din,din);}
\end{verbatim}
\end{codesize}
\caption{CUDA code for our large sorter. {\tt sortSmall} and {\tt bmergeSmall}
are replaced by Obsidian-generated kernels in the experiments. {\tt vSwap} and {\tt iSwap} are handwritten CUDA kernels
that perform one column of compare and swap operations in the vee and ilv shapes respectively, and are
parameterised on the stride. (Our generated kernels have fixed input and output size.)}
\label{fig:CUDAsort}
\end{figure}
We have measured the performance of generated 512-input sorting and merging kernels by plugging them into a larger sorter written in CUDA.
The sorter has exactly the structure shown in Figure~\ref{fig:mixedsorter}.
Figure~\ref{fig:newmixedsorter} shows the location of smaller sorters and mergers, and of vSwap and iSwap kernels for $16$ inputs.
Larger sorters simply have more columns of mergers, each preceded by a vSwap and a number of iSwaps.
The overall structure of the resulting CUDA code is shown in Figure~\ref{fig:CUDAsort}.
Because {\tt iSwap} is used repeatedly, we also wrote kernels
corresponding to compositions of two and three of them (avoiding
memory accesses between the columns).
That is, we replaced the loop containing {\tt iSwap} above by
% \pagebreak
\begin{codesize}
\begin{verbatim}
for(int j = i-1; j >= 0; j -= 3){ 
  if (j==0) 
    iSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
    else 
    {if (j==1) 
       iSwap2<<<blocks/4,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
       else 
       iSwap3<<<blocks/8,threads*2,0>>>(din,din,(1<<j)*S_SIZE);}}
\end{verbatim}
\end{codesize}
\noindent
The Obsidian implementation and all code for examples in the paper are available at \url{http://www.cse.chalmers.se/~joels/expressive.html}.

### 2013-10-06 src2/expressive/discussion.tex
hacking

The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
description language~\cite{LAVASORTER} was that a relatively small set of combinators went a long way. So, although we introduced three
combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which
includes the other two, we do not believe that every new kernel development exercise would demand a completely new set of combinators.
We expect to provide the user with a well-documented set of combinators,
so that users can get access to this style of programming without having
to develop their own combinators, and without having to think too much
about bit-hacking.
The bit-manipulation approach chosen to define our combinators automatically
created functions that apply to sub-sequences of the input that
are of an appropriate length.

>>>>

The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
description language~\citeexp{LAVASORTER} was that a relatively small set of combinators went a long way. So, although we introduced three
combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which
includes the other two, we do not believe that every new kernel development exercise would demand a completely new set of combinators.
We expect to provide the user with a well-documented set of combinators,
so that users can get access to this style of programming without having
to develop their own combinators, and without having to think too much
about bit-hacking.
The bit-manipulation approach chosen to define our combinators automatically
created functions that apply to sub-sequences of the input that
are of an appropriate length.

### 2013-10-06 src2/expressive/intro.tex
apa
hacking

\subsubsection{Embedded DSLs for GPGPU programming} 
We are aiming for a GPU programming language that is more
concise than mainstream
languages such as CUDA and OpenCL. 
Obsidian is a
domain specific embedded language (DSEL) implemented in Haskell.
When an Obsidian program is run, a representation of the program is created 
as a syntax tree. For more information on EDSL implementation see~\cite{COMPILEEDSL}.
The program representation generated when running an Obsidian program 
is compiled into CUDA code. We are also working on an OpenCL backend.

Our approach is different from that of other Haskell DSELs targetting
GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. 
We do not try to abstract away from 
all the peculiarities of GPU programming, but rather provide a higher 
level language in which to experiment with them.
For instance,
Accelerate provides a standard set of basic operations such as 
{\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATE}. 
Obsidian, on the other hand, allows the user to experiment with the
{\em generation} of small kernels for fixed size array inputs
from higher level descriptions.
It is intended to allow the user to play with the kinds of tradeoffs that are important
when writing such high performance building blocks; in this paper, the main consideration
is the number of array elements of the input and output that are manipulated
by a single thread in the generated CUDA code.
An important aspect of Obsidian is the symbolic array representation used, along
with its associated {\tt sync} operation. As we shall see, the {\tt sync} operation
allows the programmer to guide code generation and control parallelism and thread
use~\cite{JSLIC}.

%Today there are number of EDSLs written in Haskell targeting 
%GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. We justify our work on yet 
%another by having a slightly different approach. We do not try to abstract away from 
%all the peculiarities of GPU programming but rather giving a higher 
%level language in which to experiment with them.
%trying to appeal to a slightly different target audience. 
%We try to cater not for the advanced functional programmer who wants to 
%run his programs on GPUs, but rather to offer a higher level 
%language to the CUDA/OpenCL hacker. 
%In this language we want the user to be 
%able to quickly experiment with the kind of tradeoffs that are important 
%when writing a high performance GPU kernel.

%Unlike the other approaches that are complementary to our Haskell GPU EDSL, 
%Obsidian does not provide the usual set of basic operations such as 
%{\tt Map}, {\tt Reduce} and {\tt ZipWith} as built in skeletons. We 
%want a language where the programmer can experiment with implementing 
%his or her own versions of these basic operations. So far, we make
%this possible in Obsidian through a {\tt Sync} operation and the Array 
%representation we use. The {\tt Sync} operation, using target domain 
%nomenclature, is the programmers tool to guide code generation and 
%obtain parallelism. The presentation of Obsidian here will be somewhat 
%cursory, for a more complete view of the various versions and approaches 
%we have tried, see~\cite{JSLIC}.
%More details on this later.

@@

An array is represented by an indexing function and a length:
\begin{codesize} 
\begin{verbatim}
data Array a = Array (UWordE -> a) Word32 
\end{verbatim}
\end{codesize}
This array representation has served us well. It has these properties:
\begin{itemize} 
\item Fusion of operations is automatic.
\item It naturally describes a data-parallel computation suitable for CUDA/OpenCL generation.
\item Many basic operations can be implemented: {\tt map}, {\tt zipWith} etc.
\end{itemize}
Using this array representation in a DSEL is not new; the first occurence that we know of is in
Pan~\cite{PAN}. Similar array representations have also later been 
used in Feldspar~\cite{FELDSPAR}, and more recently also in the Repa library~\cite{REPA}.
Functions for indexing and getting the length of arrays are as follows:
\begin{codesize} 
\begin{verbatim}
(!) :: Array a -> UWordE -> a 
(Array ixf _) ! ix = ixf ix

@@

halve arr = splitAt ((len arr) `div` 2) arr
\end{verbatim}
\end{codesize}
A reduction kernel that takes an array whose length is a power of two 
and gives an array of length one can be defined recursively. 
Defining kernels recursively results in completely unrolled CUDA kernels,
and kernel input size must be known at compile time.
The approach to reduction taken here is to split the input array 
into two halves and then apply {\tt zipWith} of the combining function to the 
two halves, repeating the process until the length is one. 
\begin{codesize} 
\begin{verbatim}
reduceS :: (a -> a -> a) -> Array a -> Kernel (Array a) 
reduceS op arr | len arr == 1 = return arr
               | otherwise    = 
                 (pure ((uncurry (zipWith op)) . halve)
                  ->- reduceS op) arr
\end{verbatim}
\end{codesize}
%This Obsidian Kernel takes an array of length $2n$ as input and the output 
Since the output of this kernel is of length one, and the number 
of elements in the output array specifies the number of threads used to compute it, 
this function, {\tt reduceS}, defines a sequential reduction. 
The generated code for arrays of length eight is 
\pagebreak
%The code is generated with the 
%command {\tt genKernel ``reduceSAdd'' (reduce (+)) (namedArray undefined 8 :: Array IntE)}, 
%meaning that a CUDA reduce-with-addition kernel for arrays of length eight is generated. 
\begin{codesize} 
\begin{verbatim}
__global__ void reduceSAdd(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

@@

%For a more in depth view of the parts of Obsidian that has been showed 
%in this section see~\cite{JSLIC}.

>>>>

\subsubsection{Embedded DSLs for GPGPU programming} 
We are aiming for a GPU programming language that is more
concise than mainstream
languages such as CUDA and OpenCL. 
Obsidian is a
domain specific embedded language (DSEL) implemented in Haskell.
When an Obsidian program is run, a representation of the program is created 
as a syntax tree. For more information on EDSL implementation see~\citeexp{COMPILEEDSL}.
The program representation generated when running an Obsidian program 
is compiled into CUDA code. We are also working on an OpenCL backend.

Our approach is different from that of other Haskell DSELs targetting
GPUs~\citeexp{ACCELERATEDAMP11,NIKOLA,BARRACUDA}. 
We do not try to abstract away from 
all the peculiarities of GPU programming, but rather provide a higher 
level language in which to experiment with them.
For instance,
Accelerate provides a standard set of basic operations such as 
{\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
with the help of small, predefined, hand-tuned CUDA kernels~\citeexp{ACCELERATEDAMP11}. 
Obsidian, on the other hand, allows the user to experiment with the
{\em generation} of small kernels for fixed size array inputs
from higher level descriptions.
It is intended to allow the user to play with the kinds of tradeoffs that are important
when writing such high performance building blocks; in this paper, the main consideration
is the number of array elements of the input and output that are manipulated
by a single thread in the generated CUDA code.
An important aspect of Obsidian is the symbolic array representation used, along
with its associated {\tt sync} operation. As we shall see, the {\tt sync} operation
allows the programmer to guide code generation and control parallelism and thread
use~\citeexp{JSLIC}.

%Today there are number of EDSLs written in Haskell targeting 
%GPUs~\citeexp{ACCELERATE,NIKOLA,BARRACUDA}. We justify our work on yet 
%another by having a slightly different approach. We do not try to abstract away from 
%all the peculiarities of GPU programming but rather giving a higher 
%level language in which to experiment with them.
%trying to appeal to a slightly different target audience. 
%We try to cater not for the advanced functional programmer who wants to 
%run his programs on GPUs, but rather to offer a higher level 
%language to the CUDA/OpenCL hacker. 
%In this language we want the user to be 
%able to quickly experiment with the kind of tradeoffs that are important 
%when writing a high performance GPU kernel.

%Unlike the other approaches that are complementary to our Haskell GPU EDSL, 
%Obsidian does not provide the usual set of basic operations such as 
%{\tt Map}, {\tt Reduce} and {\tt ZipWith} as built in skeletons. We 
%want a language where the programmer can experiment with implementing 
%his or her own versions of these basic operations. So far, we make
%this possible in Obsidian through a {\tt Sync} operation and the Array 
%representation we use. The {\tt Sync} operation, using target domain 
%nomenclature, is the programmers tool to guide code generation and 
%obtain parallelism. The presentation of Obsidian here will be somewhat 
%cursory, for a more complete view of the various versions and approaches 
%we have tried, see~\citeexp{JSLIC}.
%More details on this later.

@@

An array is represented by an indexing function and a length:
\begin{codesize} 
\begin{verbatim}
data Array a = Array (UWordE -> a) Word32 
\end{verbatim}
\end{codesize}
This array representation has served us well. It has these properties:
\begin{itemize} 
\item Fusion of operations is automatic.
\item It naturally describes a data-parallel computation suitable for CUDA/OpenCL generation.
\item Many basic operations can be implemented: {\tt map}, {\tt zipWith} etc.
\end{itemize}
Using this array representation in a DSEL is not new; the first occurence that we know of is in
Pan~\citeexp{PAN}. Similar array representations have also later been 
used in Feldspar~\citeexp{FELDSPAR}, and more recently also in the Repa library~\citeexp{REPA}.
Functions for indexing and getting the length of arrays are as follows:
\begin{codesize} 
\begin{verbatim}
(!) :: Array a -> UWordE -> a 
(Array ixf _) ! ix = ixf ix

@@

halve arr = splitAt ((len arr) `div` 2) arr
\end{verbatim}
\end{codesize}
A reduction kernel that takes an array whose length is a power of two 
and gives an array of length one can be defined recursively. 
Defining kernels recursively results in completely unrolled CUDA kernels,
and kernel input size must be known at compile time.
The approach to reduction taken here is to split the input array 
into two halves and then apply {\tt zipWith} of the combining function to the 
two halves, repeating the process until the length is one. 
\begin{codesize} 
\begin{verbatim}
reduceS :: (a -> a -> a) -> Array a -> Kernel (Array a) 
reduceS op arr | len arr == 1 = return arr
               | otherwise    = 
                 (pure ((uncurry (zipWith op)) . halve)
                  ->- reduceS op) arr
\end{verbatim}
\end{codesize}
%This Obsidian Kernel takes an array of length $2n$ as input and the output 
Since the output of this kernel is of length one, and the number 
of elements in the output array specifies the number of threads used to compute it, 
this function, {\tt reduceS}, defines a sequential reduction. 
The generated code for arrays of length eight is 
%\pagebreak
%The code is generated with the 
%command {\tt genKernel ``reduceSAdd'' (reduce (+)) (namedArray undefined 8 :: Array IntE)}, 
%meaning that a CUDA reduce-with-addition kernel for arrays of length eight is generated. 
\begin{codesize} 
\begin{verbatim}
__global__ void reduceSAdd(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

@@

%For a more in depth view of the parts of Obsidian that has been showed 
%in this section see~\citeexp{JSLIC}.

### 2013-10-06 src2/expressive/push.tex
apa
hacking

\section{Push Arrays}

@@

\subsection {What are Push Arrays?}

@@

\noindent
For another example of using continuations and a more complete
description of their meaning and application, see~\cite{POORKOEN}.

@@

Note that the {\tt Par} constructor, 
the {\em parallel for loop}, could potentially introduce nesting, which would lead 
to {\em nested data-parallelism}. We do not compile nested data parallelism 
into CUDA, and right now this is guaranteed by taking care not to introduce 
any nesting in the library functions provided. Some of the 
simpler cases of nestedness should be possible to take care of quite easily. 
For example, one extra level of nesting could be done by sequential execution 
in each thread of the GPU; using sequential computations per thread has 
been shown to be beneficial~\cite{OLAMARCUS}. But for the general case 
of arbitrary nesting, some method of flattening is needed. 
We also assume that both {\tt Allocate} and {\tt Synchronize} occur only at the top level in objects of type {\tt Program}.

>>>>

\subsection{Push Arrays}

@@

\subsubsection {What are Push Arrays?}

@@

\noindent
For another example of using continuations and a more complete
description of their meaning and application, see~\citeexp{POORKOEN}.

@@

Note that the {\tt Par} constructor, 
the {\em parallel for loop}, could potentially introduce nesting, which would lead 
to {\em nested data-parallelism}. We do not compile nested data parallelism 
into CUDA, and right now this is guaranteed by taking care not to introduce 
any nesting in the library functions provided. Some of the 
simpler cases of nestedness should be possible to take care of quite easily. 
For example, one extra level of nesting could be done by sequential execution 
in each thread of the GPU; using sequential computations per thread has 
been shown to be beneficial~\citeexp{OLAMARCUS}. But for the general case 
of arbitrary nesting, some method of flattening is needed. 
We also assume that both {\tt Allocate} and {\tt Synchronize} occur only at the top level in objects of type {\tt Program}.
